Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-12-09 manuscript.Rmd
Revisions for review response.

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface. Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant. Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28. Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling. High-touch surfaces (Table 1) were selected using structured observations prior to study start. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction (QIAmp Viral RNA Mini Kit) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did. The highest probability of contamination was found with floors, followed by elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory support. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. Our findings match previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets or aerosols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Second, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning. Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

@@

dat %>%
  left_join(distinct(select(dat_long, swab_site, site_descriptor)), by = "swab_site") %>%
  mutate(high_touch = case_when(high_touch == TRUE ~ "High Touch",
                                high_touch == FALSE ~ "Low Touch")) %>%
  mutate(site_descriptor = stringr::str_to_title(string = gsub("_"," ",gsub("1|2","",site_descriptor)))) %>%
  group_by(site_category, site_descriptor, high_touch) %>%
  summarise(unique_subjects = n_distinct(subject_room_id),
            unique_specimens = n(),
            prop_positive = sum(!is.na(copies_max)) / n(),
            med_copies = median(log10(copies_max), na.rm = TRUE),
            q25_copies = quantile(log10(copies_max), probs = 0.25, na.rm = TRUE),
            q75_copies = median(log10(copies_max),  probs = 0.75, na.rm = TRUE),
            copies_summary = glue("{round(med_copies,1)} ({round(q25_copies,1)} - {round(q75_copies,1)})")) %>%
  ungroup() %>%
  arrange(desc(prop_positive)) %>%
  select(- contains("_copies")) %>%
  gt::gt() %>%
  gt::cols_label(site_category = "Height", site_descriptor = "Location", high_touch = html("Contact"), unique_subjects = html("Subjects<br>Sampled"), unique_specimens = html("Surfaces<br>Sampled"), prop_positive = html("SARS-CoV-2<br>Detected"), copies_summary = html("Median (IQR) log<sub>10</sub><br>Copies by RT-qPCR")) %>%
  gt::fmt_percent(columns = prop_positive, decimals = 1) %>%
  identity() -> dat_gt
dat_gt

@@

__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are reproted.

>>>>

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol <mark>(**Supplemental Methods**). Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface. Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.</mark> Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28. <mark>Sampling was performed each morning prior to daily cleaning.</mark> Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling. High-touch surfaces (Table 1) were selected using structured observations prior to study start. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction (QIAmp Viral RNA Mini Kit) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces <mark>(mouse, keyboard, bed rail, and doorknob)</mark> had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces <mark>(walls)</mark>. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did. The highest probability of contamination was found with floors, followed by elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or <mark>other mobile medical equipment</mark> increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory support. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@Cdc2021-du].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. Our findings match previous description of high probability of floor contamination relative to other surfaces, <mark>but the role of mops in the spread of viral RNA has not been previously described [@Chia2020; @Rutala2007-ki]</mark>. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets or aerosols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. <mark>Few prior studies that have been successful in recovering infectious virus from patient rooms [@Lednicky2020-ue; @Santarpia2020-ll].</mark> Second, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning. <mark>Third, we were not able to differentiate contamination of floors that was attributable to healthcare personnel foot traffic as opposed to the settling of respiratory particles or body fluids attributable to the index patient.  We did not observe large differences in the abundance of SARS-CoV-2 RNA across surfaces where SARS-CoV-2 RNA was detected. Therefore, our analysis focused exclusively on factors that are associated with the presence or absence of SARS-CoV-2, rather than SARS-CoV-2 abundance on positive surfaces.</mark> Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

@@

dat %>%
  left_join(distinct(select(dat_long, swab_site, site_descriptor)), by = "swab_site") %>%
  mutate(high_touch = case_when(high_touch == TRUE ~ "High Touch",
                                high_touch == FALSE ~ "Low Touch")) %>%
  mutate(site_descriptor = stringr::str_to_title(string = gsub("_"," ",gsub("1|2","",site_descriptor)))) %>%
  group_by(site_category, site_descriptor, high_touch) %>%
  summarise(unique_subjects = n_distinct(subject_room_id),
            unique_specimens = n(),
            prop_positive = sum(!is.na(copies_max)) / n(),
            med_copies = median(log10(copies_max), na.rm = TRUE),
            q25_copies = quantile(log10(copies_max), probs = 0.25, na.rm = TRUE),
            q75_copies = median(log10(copies_max),  probs = 0.75, na.rm = TRUE),
            copies_summary = glue("{round(med_copies,1)} ({round(q25_copies,1)} - {round(q75_copies,1)})")) %>%
  ungroup() %>%
  arrange(desc(prop_positive)) %>%
  select(- contains("_copies")) %>%
  gt::gt() %>%
  gt::cols_label(site_category = "Height", site_descriptor = "Location", high_touch = html("Contact"), unique_subjects = html("Subjects<br>Sampled"), unique_specimens = html("Surfaces<br>Sampled"), prop_positive = html("SARS-CoV-2<br>Detected"), copies_summary = html("Median (IQR) log<sub>10</sub><br>Copies by RT-qPCR")) %>%
  gt::fmt_percent(columns = prop_positive, decimals = 1) %>%
  #gt::tab_footnote()
  identity() -> dat_gt
dat_gt

@@

__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are <mark>reported. The "exit" descriptor indicates sites within the patient room near the exit or threshold; the "near" descriptor indicates sites within the patient room near the patient bed</mark>.
